\Large If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-12-05 GroupPolarizationStatmod/paper/discussion.tex
starting penultimate revision

>>>>

\subsection{The Full Bayesian Rigor: Deriving Effect Size Posteriors and Justifying the FPR Approximation}

In evaluating the \textbf{False Positive Rate (FPR)} of a simulated null model, the primary challenge is moving beyond the raw parameter posterior distributions ($p(\mu_1 | \text{data})$, $p(\sigma_2 | \text{data})$, etc.) to accurately quantify the posterior distribution of the derived \textbf{effect size}, $d$. This section outlines the statistically rigorous method for this derivation and justifies the pragmatic approximation necessary for the evaluation stage of this paper.

\subsubsection{The most rigorous formulation of $d$}

In our model, the effect size $d$ is a function of four estimated parameters ($\mu_1, \mu_2, \sigma_1, \sigma_2$), defined by the authors whose work we replicate as:

\begin{equation}
d = \frac{\mu_1 - \mu_2}{\sqrt{(\sigma_1^2 + \sigma_2^2)/2}}
\end{equation}

The statistically rigorous method for obtaining the posterior distribution of $d$, denoted $p(d | \text{data})$, is through \textbf{marginalization by sampling} across the joint posterior distribution of the primary parameters $p(\mu_1, \mu_2, \sigma_1, \sigma_2 | \text{data})$.

\paragraph{Necessity of MCMC Samples}

The parameters are fundamentally \textbf{sequentially dependent} in the MCMC chain, as the estimation of means is conditional on the estimation of variances. This dependency is captured in the \textbf{joint posterior distribution}.

To maintain statistical rigor, we must avoid calculating $d$ by combining the isolated summary statistics (e.g., the posterior median of $\mu_1$ and the posterior median of $\sigma_1$) because this procedure, known as \textbf{plug-in estimation}, ignores the critical correlation between parameters, resulting in an unquantified and likely underestimated uncertainty.

The correct procedure requires using the full MCMC samples:

\begin{enumerate}
    \item For each usable iteration $k$ of the Markov Chain (where $K$ is the total number of samples), a plausible set of values is drawn: $(\mu_1^{(k)}, \mu_2^{(k)}, \sigma_1^{(k)}, \sigma_2^{(k)})$.
    \item The effect size is calculated for that specific set of values:
    $$d^{(k)} = \frac{\mu_1^{(k)} - \mu_2^{(k)}}{\sqrt{\frac{(\sigma_1^{(k)})^2 + (\sigma_2^{(k)})^2}{2}}}$$
    \item The set $\{d^{(1)}, d^{(2)}, \ldots, d^{(K)}\}$ then forms the complete posterior distribution $p(d | \text{data})$.
\end{enumerate}

The center of this distribution, the \textbf{posterior median} ($\tilde{d}$), is the most robust point estimate for the effect size, as it minimizes expected absolute error and is less sensitive to potential skewness than the posterior mean.

\subsubsection{Justifying the FPR Approximation}

Due to the computational complexity and the unavailability of the original MCMC chains for all $57,000$ fits, a pragmatic simplification was necessary for the subsequent evaluation of the False Positive Rate (FPR).

\paragraph{The FPR Criterion}

A standard method for estimating the FPR (or Type I error rate) in simulation studies is to count the proportion of null trials where the observed effect size meets or exceeds a predefined threshold, $d^*$.

$$\text{FPR} = \frac{\text{Number of Null Fits where } |d| > d^*_{\text{threshold}}}{\text{Total Number of Null Simulations}}$$

In this analysis, $d^*_{\text{threshold}}$ is set to the commonly adopted small effect size threshold of $d^*=0.2$ (and varied for sensitivity analysis), avoiding the reliance on approximated $95\%$ credible intervals, which would require further, less reliable approximations.

\paragraph{The Compromise: Plug-in and Heuristics}

We employed a simplified plug-in value for $d$ using the stored posterior means:

$$d_{\text{plug}} = \frac{\hat{\mu}_1 - \hat{\mu}_2}{\sqrt{(\hat{\sigma}_1^2 + \hat{\sigma}_2^2)/2}}$$

Where $\hat{\theta}$ represents the \textbf{posterior mean} of the parameter $\theta$.

While this approach is a statistical shortcut, the error is mitigated by two factors: the use of the $d^*_{\text{threshold}}$ heuristic, and the large simulated sample size ($N=500$ per group), which ensures that the posteriors for $\mu$ and $\sigma$ are sufficiently narrow. This minimizes the difference between $d_{\text{plug}}$ and the true posterior median $\tilde{d}$, making the shortcut defensible as a pragmatic necessity given the study's scale.
